The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-16 tasbot/paper/paper.tex
More work on the paper. Getting closer...

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
% \usepackage{inlinebib}

@@

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  {\bf Parameter Alert!}\,\, #1
}

@@

\begin{abstract}
This is where the Abstract goes.
\end{abstract}

@@

\section*{Introduction}
The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed. Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites. By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware. This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games. The approach is practical on a single computer, and succeeds on several games, such as Super Mario Bros.. The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}

@@

% (32kB of ROM is addressable at a time, and most games used bank switching to access hundreds of kilobytes of game included in the cartridge.)

There are a number of emulators for NES. These work by simulating the NES hardware, for example with a 2048-byte array for its memory, and simulating the steps of its 2A03 processor on some ROM, and hooking a keyboard or joystick into the 8 bits of input. (There are of course many details to work out! But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 bytes per second) and play them back and get the same result. This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run. In this work we use the FCEUX emulator, which is popular for its accuracy and advanced tools.

@@

@@

{\bf Skipping.}\,\, To avoid being confused by RAM initialization and
menu, I ignore all memories up until the first input by the player.
Including these would be especially suspicious because the RAM's
contents are not even well-defined until the program writes something
to them.\footnote{Emulators tend to fix them to a specific pattern so
  that emulation is completely deterministic, but in real hardware
  they are truly uninitialized, or retain the values from the last
  reset or game inserted. Some semi-famous tricks involve removing
  and inserting game cartridges while the system is running in order
  to take advantage of this.}

% XXXX good place for some graphs.
{\bf Slicing.}\,\, I generate 50 orderings for $M_1 \ldots M_n$; the
whole recording starting from the first keypress. During gameplay some
values really are completely nondecreasing, like the player's score
and world/level pair. I also generate 3 orderings for each tenth of
the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and $M_{n/10+1} \ldots
M_{2n/10}$, etc. The intention is to capture orderings that are rarely
violated, or sections of the game with unusual objectives (e.g. a
minigame or swimming level). Orderings generated this way look pretty
random, and on a scale from solid to suspicious, I can't vouch for
them. Then I generate objectives from non-consecutive memories that
are evenly spread out through the observations: Ten objectives
chosen from every 100\th\ memory, starting from the 0\th\ frame,
1\st\ frame, 2\nd\ frame, and so on up to the 9\th{}. Similarly
for every 250\th\ frame, and a single objective for memory sliced
to every 1000\th\ frame, with start positions of 0--9.

{\bf Weighting.},\, To reduce the importance of randomness in the
learning process, and the arbitrariness of the slicing, each objective
function is also assigned a weight. An ideal objective function takes
on its minimal value on the first frame and maximal on the last
(according to the ordering), and increases steadily throughout the
observed memories. This is ideal because it allows us to just follow
the gradient to reach good states. A bad objective function freqently
regresses (recall that although we produce valid orderings, an
ordering that is valid on some slice may not be valid for the whole
sequence of memories). To weight an objective $L$, we first collect
all of the values (the vector of values of the memory locations $L_1
\ldots L_k$) it takes on throughout the observations. These may not
obey the ordering. We then sort the values lexicographically and
remove duplicates.\footnote{Note to self: I'm not sure how to justify
  removing duplicates here. It makes [0, 1, 1, 1, 1, 1, 10, 11] look
  the same as [0, 1, 10, 11], which is probably not desirable?} Call
this $V$. Now we can compute the {\em value fraction} for $L$ on some
$M$: Extract the vector of locations $M[L_1], M[L_2], \ldots, M[L_k]$
and find the lowest index $i$ in $V$ where the vector is less than or
equal to $V_i$. The value fraction {\tt VF} is $i/|V|$, which is the
normalized value of ``how big'' $M$ is, according to $L$, compared to
all the values we ever saw for it. The value fraction is defined and
in $[0, 1]$ for all memories in the observed set.\footnote{It is not
  defined when the memory is greater than all observed memories, which
  we will encounter later. The code returns $(i+1)/|V|$ in that case,
  which is as good as anything.} This gives us the ability to compare
objectives on an absolute scale.\footnote{This is certainly not the
  only way to do it, and it has some questionable properties like
  ignoring the magnitude of change. But it is very simple.} Weighting
an objective is now simple:
%
$$ \Sigma_{i=1}^{n-1}  {\tt VF}(M_{i+1}) - {\tt VF}(M_{i}) $$
%
We sum the differences in value functions for each consecutive pair of
memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
1/n$, which approaches $1$.\footnote{The code implements a different
  weighting scheme, which sums from $i=0$ and keeps track of the last
  value fraction at each step, thus treating the value fraction of the
  nonexistent memory $M_0$ as 0. This is wrong, because the first
  value fraction may be very high, which credits the objective with a
  positive value (e.g. 1) for that first part of the sum. Objectives
  that start high on the first frame are not ideal; in fact, the worst
  objectives start high on the first frame and steadily {\em
    decrease}. I plan to correct this bug and study if it has any
  noticeable consequences.} Objectives are constrained to have
non-negative weights (I set the value to 0 if negative, which
effectively disables it).

% XXX good place to give some weights for some objectives.

We save the objectives and their weights to a file and then we are
done with the easy part.

@@

\section{Now you're playing with power}

In this section I'll describe how the objective functions are used to
play the game. On a scale from canonical to Star Wars Christmas
Special, this algorithm is an 7. So, rather than focus on the
particulars of some of the heuristics, I'll try to give a story of the
different things I tried, what motivated the ideas in the current
version, and my intuitions about what is working well (or not) and
why. This algorithm is called {\tt playfun} and it can be found
implemented in C++ in {\tt playfun.cc}.

@@

In order to use the emulator to search for good sequences of inputs, I
needed deeper integration than just observing memory. The FCEUX
emulator is about a jillion lines of C++-ish code, was intended as an
interactive GUI application, contains support for multiple different
platforms, and the code is, on a scale from a pile of horse shit to
not horse shit, approximately a 2.\footnote{It is, however, an
  excellent emulator to use, has fancy tools for recording and editing
  movies, and is popular in the speedrun community; just don't read
  the code.} With a medium amount of suffering I got the emulator
compiling under mingw in 64-bit mode, and working behind a streamlined
interface ({\tt emulator.h}). Once a game is initialized, it is always
at an input frame---you can give it an 8-bit input (for the
1\st\ player) to step a single frame, or read the 2048 bytes of
memory. You can also save the complete state of the emulator into a
vector of bytes, which allows you to restore the state to exactly that
same point.\footnote{This contains the RAM, but also stuff we didn't
  consider, like the Picture Processing Unit's state, and internal
  emulator stuff.} These save-states are portable across sessions as
long as the code was compiled and the game initialized the right
way.\footnote{The original FCEUX supports portable save-states, but I
  removed that guarantee in order to get the smallest possible byte
  vectors. More on that below.} FCEUX must be single-threaded because
it uses global variables galore. I made several enhancements to the
emulator interface, which are discussed later.

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames;
it takes about XXX ms to process a single step. Lots of engineering
effort goes into reducing the number of frames the algorithms emulate.

\subsection{Naive attempts}

@@

Taking the minimum of random futures is always silly (at least if we
believe our objective function), because nothing other than our own
bad memory can force us to take a series of steps if we know a
different better series of steps. Taking the max is possibly foolish
if we don't know how to reconstruct a rare good state that caused the
maximum score to be high. Both lead us to the idea: Keep around
a set of candidate futures that we use for evaluation and search,
rather than just randomly generating futures when we want to evaluate
a state. This turns out to work really well and be more efficient.

HERE

\subsection{Caching of emulation}

% good parts!
% pacman @6800

\section{Results}

@@

@@

% ends with 13 dots remaining

@@

\subsection{Bubble Bobble}

@@

I was surprised that automating this game works at all, since there is
no obvious spatial gradient like in Super Mario Bros., and few local
greedy rewards like in Pac-Man. Bub initially wastes his lives (a
common theme in games where the respawn interval is low---it just
looks like a fast way of warping to the corner). When he's near
monsters he shows good tactics in shooting and popping them
simultaneously, even doing so while facing away from them, which I
didn't know was possible. At the beginning of the movie he prefers to
hide in the bottom-left corner, but soon starts jumping all around the
level, bouncing on his own bubbles and so on, almost like a real
player would. He beats XXX levels, more than I would have guessed.
Since it takes two start button presses to enter the game, the second
START is part of the input motifs. Amusingly, Bub uses this to his
advantage to change the synchronization of his futures, and when the
situation gets really tight, he'll pause the game until a good
future comes around (Figure~\ref{fig:bubblebobble}). Paused states
probably look like modest progress since memory locations like the
music cursor are still advancing.

XXX what finally happens?

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}
% \usepackage{inlinebib}

@@

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  \begin{quotation}
  {\bf Parameter Alert!}\,\, #1
  \end{quotation}
}

@@

\begin{abstract}
% XXX better abstract
This paper presents a simple, generic method for automating the play
of Nintendo Entertainment System games.
\end{abstract}

@@

\section{Introduction}
The Nintendo Entertainment System is probably the best video game
console, citation {\it not} needed. Like many, I have spent thousands
of hours of my life playing NES games, including several complete
playthroughs of classics like Super Mario Bros., Bionic Commando,
Bubble Bobble, and other favorites. By the year 2013, home computers
have become many orders of magnitude faster and more capacious than
the NES hardware. This suggested to me that it may be time to automate
the playing of NES games, in order to save time.\footnote{Rather, to
  replace it with time spent programming.} In this paper I present a
generic technique for automating the playing of NES games. The
approach is practical on a single computer, and succeeds on several
games, such as Super Mario Bros.. The approach is amusingly elegant
and surprisingly effective, requires no detailed knowledge of the game
being played, and is capable of novel and impressive gameplay (for
example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience:
  This work is 100\% real.}

@@

% (32kB of ROM is addressable at a time, and most games used bank
% switching to access hundreds of kilobytes of game included in the
% cartridge.)

There are a number of emulators for NES. These work by simulating the
NES hardware, for example with a 2048-byte array for its memory, and
simulating the steps of its 2A03 processor on some ROM, and hooking a
keyboard or joystick into the 8 bits of input. (There are of course
many details to work out! But in essence emulation is just that.) This
process is completely deterministic, so it is possible to record the
sequence of inputs (the inputs can only be read once per video frame,
so this sequence is 60 bytes per second) and play them back and get
the same result. This also means that an input sequence can be
computed in non-real time, either much slower or much faster than a
NES would normally run. In this work we use the FCEUX\cite{FCEUX}
emulator, which is popular for its accuracy and advanced tools.

@@

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-global}
\end{center}\vspace{-0.1in}
\caption[lof not used]{
  A single maximal tight valid lexicographic ordering for my
  4,000-frame input training data to Super Mario Bros.. This function
  is globally nondecreasing, and is the decimal memory locations
  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.

  This is not a great objective function; there are long spans where
  all the memories are equal according to it, and the nice smooth slopes
  are happening during level transitions where the game is ignoring
  inputs (they are probably timers counting up each frame, which is
  why they are so smooth). Other slicing produces better objectives.

  For reasons unknown---I just discovered this while generating the
  figure---{\em all} of the objective functions learned with this
  method, regardless of the nondeterministic choices, appear to have
  this same curve, despite using different memory locations. It may be
  that they are different permutations of bytes that all change
  simultaneously, only on the frames where there are jumps in this
  picture, and there are no other orderings that are tight, valid, and
  maximal. This is still surprising and warrants investigation. }
\label{fig:global}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-tenths}
\end{center}\vspace{-0.1in}
\caption{Ten objective functions trained on different tenths of the
  4,000 inputs for Super Mario Bros.. These functions are normalized
  against the range of all values they take on during the movie; you
  can see that most are increasing locally most of the time, but many
  drop back to zero around the 3100\th\ frame, when Mario reaches
  world {\tt 1-2}. Within its 400-frame window, each objective is
  guaranteed to be nondecreasing.}
\label{fig:tenths}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-strides}
\end{center}\vspace{-0.1in}
\caption{Ten objective functions trained on every 100\th\ memory,
  starting at frame 0, frame 1, and so on up to frame 10. Normalized
  as usual. These objectives exhibit excellent global progress, but
  are locally very noisy. Among the methods I used to generate
  objectives, this one produces the best results (based on my
  intuition about what a good objective function looks like).
}
\label{fig:strides}
\end{figure}

@@

\paragraph{Skipping.} To avoid being confused by RAM initialization and
menu, I ignore all memories up until the first input by the player.
Including these would be especially suspicious because the RAM's
contents are not even well-defined until the program writes something
to them.\footnote{Emulators tend to fix them to a specific pattern so
  that emulation is completely deterministic, but in real hardware
  they are truly uninitialized, or retain the values from the last
  reset or game inserted. Some semi-famous tricks involve removing and
  inserting game cartridges while the system is running in order to
  take advantage of this.}

\paragraph{Slicing.} I generate 50 orderings for $M_1 \ldots M_n$; the
whole recording starting immediately after the first keypress. During
gameplay some values really are completely nondecreasing, like the
player's score and world/level pair. Figure~\ref{fig:global} shows
what a global ordering looks like. I also generate 3 orderings for
each tenth of the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and
$M_{n/10+1} \ldots M_{2n/10}$, etc. The intention is to capture
orderings that are rarely violated, or sections of the game with
unusual objectives (e.g. a minigame or swimming level). Orderings
generated this way look pretty random, and on a scale from solid to
suspicious, I can't vouch for them. Then I generate objectives from
non-consecutive memories that are evenly spread out through the
observations: Ten objectives chosen from every 100\th\ memory,
starting from the 0\th\ frame, 1\st\ frame, 2\nd\ frame, and so on up
to the 9\th{}. Similarly for every 250\th\ frame, and a single
objective for memory sliced to every 1000\th\ frame, with start
positions of 0--9. The intention is to capture objectives that grow
slowly over the course of a playthrough, without getting distracted by
local noise.

\paragraph{Weighting.} \label{sec:objectiveweighting} 
To reduce the importance of randomness in the
learning process, and the arbitrariness of the slicing, each objective
function is also assigned a weight. An ideal objective function takes
on its minimal value on the first frame and maximal on the last
(according to the ordering), and increases steadily throughout the
observed memories. This is ideal because it allows us to just follow
the gradient to reach good states. A bad objective function freqently
regresses (recall that although we produce valid orderings, an
ordering that is valid on some slice may not be valid for the whole
sequence of memories). To weight an objective $L$, we first collect
all of the values (the vector of values of the memory locations $L_1
\ldots L_k$) it takes on throughout the observations. These may not
obey the ordering. We then sort the values lexicographically and
remove duplicates.\footnote{Note to self: I'm not sure how to justify
  removing duplicates here. It makes [0, 1, 1, 1, 1, 1, 10, 11] look
  the same as [0, 1, 10, 11], which is probably not desirable?} Call
this $V$. Now we can compute the {\em value fraction} for $L$ on some
$M$: Extract the vector of locations $M[L_1], M[L_2], \ldots, M[L_k]$
and find the lowest index $i$ in $V$ where the vector is less than or
equal to $V_i$. The value fraction {\tt VF} is $i/|V|$, which is the
normalized value of ``how big'' $M$ is, according to $L$, compared to
all the values we ever saw for it. The value fraction is defined and
in $[0, 1)$ for all memories in the observed set.\footnote{It is not
    defined when the memory is greater than all observed memories,
    which we will encounter later. The code returns $|V|/|V| = 1$ in
    that case, which is as good as anything.} This gives us the
  ability to compare objectives on an absolute scale.\footnote{This is
    certainly not the only way to do it, and it has some questionable
    properties like ignoring the magnitude of change. But it is very
    simple.} Weighting an objective is now simple:
%
$$ \Sigma_{i=1}^{n-1}  {\tt VF}(M_{i+1}) - {\tt VF}(M_{i}) $$
%
We sum the differences in value functions for each consecutive pair of
memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
1/n$, which approaches $1$. Of course, when you think about it,
this is actually the same as
%
\[
\begin{array}{c}
({\tt VF}(M_1) - {\tt VF}(M_0)) + ({\tt VF}(M_2) - {\tt VF}(M_1)) \,+ \\
\vdots \\
({\tt VF}(M_{m-1}) - {\tt VF}(M_{m-2})) + ({\tt VF}(M_{m}) - {\tt VF}(M_{m}))
\end{array}
\]
%
and all but $-{\tt VF}(M_0)$ and ${\tt VF}(M_m)$ cancel out. This
means that the weight we use is just the final value minus the initial
value, regardless of what happens in-between.\footnote{I think this
  can be improved, for example by taking the deviation from the ideal
  linear objective.} The mean value theorem or something is probably
relevant here. {\bf Lesson about being careful:} I only realized that
this code was kind of fishy when I started writing it up for SIGBOVIK.
Not only did it loop over all the deltas as in the $\Sigma$ expression
above, but it also summed from $i=0$ and kept track of the last value
fraction at each step, thus treating the value fraction of the
nonexistent memory $M_0$ as 0. This is wrong, because the first value
fraction may be very high, which credits the objective with a positive
value (e.g. 1) for that first part of the sum. Objectives that start
high on the first frame are not ideal; in fact, the worst objectives
start high on the first frame and steadily {\em decrease}. After
writing this up I corrected it to the simple expression ${\tt VF}(M_m)
- {\tt VF}(M_0)$ and the result was a huge breakthrough in the
quality of the output! I had spent much time tuning the {\tt
  playfun} search procedure (Section~\ref{sec:playfun}) and not
investigated whether the objectives were being weighted properly. More
on this later, but the lesson is: Bugs matter, and thinking about your
code and explaining it is a good way to find bugs.

Objectives are constrained to have non-negative weights (I set the
value to 0 if negative, which effectively disables it). We save the
objectives and their weights to a file and then we are done with the
easy part.

@@

\section{Now you're playing with power} \label{sec:playfun}

In this section I'll describe how the objective functions are used to
play the game. On a scale from canonical to Star Wars Christmas
Special, this algorithm is an 7. So, rather than focus on the
particulars of some of the heuristics, I'll try to give a story of the
different things I tried, what motivated the ideas in the current
version, and my intuitions about what is working well (or not) and
why. This algorithm is called {\tt playfun} and it can be found
implemented in C++ in {\tt playfun.cc}; some historic versions are
in {\tt playfun-*.cc}.

@@

In order to use the emulator to search for good sequences of inputs, I
needed deeper integration than just observing memory. The FCEUX
emulator is about a jillion lines of C++-ish code, was intended as an
interactive GUI application, contains support for multiple different
platforms, and the code is, on a scale from a pile of horse shit to
not horse shit, approximately a 2.\footnote{It is, however, an
  excellent emulator to use, has fancy tools for recording and editing
  movies, and is popular in the speedrun community. I highly recommend
  it; just don't read the code.} With a medium amount of suffering I
got the emulator compiling under mingw in 64-bit mode, and working
behind a streamlined interface ({\tt emulator.h}). Once a game is
initialized, it is always at an input frame---you can give it an 8-bit
input (for the 1\st\ player) to step a single frame, or read the 2048
bytes of memory. You can also save the complete state of the emulator
into a vector of bytes, which allows you to restore the state to
exactly that same point.\footnote{This contains the RAM, but also
  stuff we didn't consider, like registers, the Picture Processing
  Unit's state, and internal emulator stuff.} These save-states are
portable across sessions as long as the code was compiled and the game
initialized the right way.\footnote{The original FCEUX supports
  portable save-states, but I removed that guarantee in order to get
  the smallest possible byte vectors. More on that below.} FCEUX must
be single-threaded because it uses global variables galore. I made
several enhancements to the emulator interface, which are discussed
later.

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames; it
takes about 500\textmu s to process a single step. Lots of engineering
effort goes into reducing the number of frames the algorithms emulate.
The {\tt playfun} program takes a ROM file, the learned objectives and
motifs, and runs on the console for arbitrarily long, outputting a
movie file consisting of the input sequence it think is best. The
current {\tt playfun} algorithm is much too slow to run real-time, but
it would be possible to have video output as it played. I disabled
most of the video code, however, in an effort to make the emulation
loop run as fast as possible.

\subsection{Naive attempts} \label{sec:naive}

@@

Taking the minimum of random futures is always silly (at least if we
believe our objective function), because nothing other than our own
bad memory can force us to take a series of steps if we know a
different better series of steps. Taking the max is possibly foolish
if we don't know how to reconstruct a rare good state that caused the
maximum score to be high. Both lead us to the idea: Keep around a set
of candidate futures that we use for evaluation and search, rather
than just randomly generating futures when we want to evaluate a state
and then discarding them. This turns out to work really well and be
more efficient.

The basic version of the {\tt playfun} algorithm looks like this.
Maintain {\tt NFUTURES} futures (this is 40 for the results presented
in Section~\ref{sec:results}), originally just seeded with weighted
random motifs. We aren't likely to execute any of these futures
verbatim, but they are intended to give us high watermarks of what
we're capable of, plus allow us to judge future danger. As we execute
search, futures will be partly consumed and extended, and some discarded
and replaced.

Each future stores a desired length from 50--800 frames, and whenever
it is shorter than that, we extend it (at its end) with random weighted
motifs. The inner pseudoloop then looks like this:

\begin{code}
for (;;) \{
  vector<uint8> before = GetMemory();
  vector<uint8> state = GetState();

 \

  set<vector<uint8>> nexts;
  for (Future f : futures) \{
   nexts.insert(f.First10Frames());
   f.ChopOffFirst10Frames();
  \}

 \

  while (nexts.size() < NFUTURES)
    nexts.push\_back(/* random motif */);

 \

  for (vector<uint8> &next : nexts) \{
    RestoreState(state);
    for (uint8 i : next) Step(i);
    double score =
      ScoreByFutures(before, futures);
    // Save the best-scoring next...
  \}

 \

  ReweightMotifs(best\_next, motifs);
  ReplaceBadFutures(futures);
  ExtendFuturesToDesiredLengths(futures);
\}
\end{code}

At each iteration of the loop we will find the best next 10-frame
sequence to commit to. Rather than search over all motifs, we search
over the first 10 frames of each future. This has several important
consequences. First, since futures are generated by weighted motifs,
it makes it more likely that we spend CPU time evaluating motifs that
are common; the code from Section~\ref{sec:naive} always explores
every motif, even rare ones. Second, it guarantees that if we pick
some 10 frames on the basis of a single good future, we don't have to
worry about recreating that future; it will still be among our futures
for the next round. This is key: It allows us to use the determinism
of the game to replay a really good future if we find one, not just
use average random futures to assess how good a state will be.

The function {\tt ScoreByFutures} saves the state, then runs each
of the {\tt NFUTURES} futures to get a final state and memory. We
score each final memory relative to the start memory. The particulars
of scoring are not as interesting as the general idea, which is:
\begin{itemize}
\item The potential 10-frame {\tt next} sequence that we're
evaluating gets an {\em optimistic} score. This is based on the futures
for which the objectives {\em go up}. This is always non-negative.
\item Each future is also given a score, which is the sum of scores
from all the different {\tt next} sequences, regardless of their sign.
\end{itemize}

The purpose of the first is to identify a good {\tt next} sequence. We
take the sequence with the highest optimistic score. The idea is that
this is the sequence that gives us the best outcome if we continue
to control what we do in the future.

The purpose of the second is to identify which futures are good. A
good future tends to bring good outcomes regardless of the immediate
next step. Random futures that make is walk left when the good stuff
is to the right, or jump when there are spikes nearby above our head,
will receive negative scores for many or all of the {\tt next}
sequences.

{\tt ReweightMotifs} changes the weight of motifs that match the best
{\tt next} sequence that we just committed to, if we think that we
made progress on this step. The idea is to learn which sequences tend
to work well for us; this might be different from what the human
player did. For example, in run-to-the-right-and-jump kinds of games,
human players tend to hesitate more before obstacles and enemies than
a computer does.\footnote{That said, this part was an early idea and
  is probably not necessary. It's suspicious because these 10-frame
  sequences are not necessarily motifs (10 is the same as the normal
  motif length, and futures are generated by motifs, but they can
  become desynchronized because of the final incomplete motif, future
  mutation, and other things). So sometimes the chosen sequence
  doesn't affect weights. I think this would be better if we kept a
  Markov model } Knowing whether we made progress is kind of
difficult, since we can almost always find some move that increases
the objectives locally. For this I use the same approach of {\em value
  fraction} from Section~\ref{sec:objectiveweighting} based on a
sample of memories that we've seen so far. If we appear to be
progressing then the motif is weighted up; if we're regressing then it
is weighted down.

{\tt ReplaceBadFutures} kicks out the the futures with the worst total
scores, so that over time the random futures become good futures. Of
course, we always have to keep randomly adding to the end of each
future, and who knows what crap we'll find? A late addition to the
algorithm replaces some proportion of these with mutated versions of
the best scoring future. This helps us from messing up a good future
by appending crap to it, since we have multiple copies of it. It also
makes search more efficient, since futures that share common prefixes
can often benefit from caching (Section~\ref{sec:caching}). Currently
mutating a future involves chopping off its second half (it will get
extended to the desired length in the next step), and a \sfrac{1}{7}
chance of {\em dualizing} the entire sequence. Dualization swaps
the left and right buttons, the up and down buttons, B and A, and 
select and start. The idea of this is to occasionally introduce
very different futures to help us get out of stuck situations where
we should really be turning around or something like that.

During the development and tuning of the algorithm, I sometimes
observed {\tt ReweightMotifs} and {\tt ReplaceBadFutures} conspiring
to create a total monoculture of futures, where the weight of a single
motif (specifically ``press no buttons'') went out of control and all
futures just consisted of waiting. Waiting is usually a local
improvement to some objectives because of internal frame counters and
the cursor used to control music playback. To guard against this,
motifs have a maximum (and minimum) possible weight in {\tt
  ReweightMotifs}, and {\tt ReplaceBadFutures} always ensure that
some fraction of the futures are completely random (ignoring motif
weights).

\parameteralert{
  This is really the worst part in terms of parameters. They are: 
  {\tt NFUTURES}, the number of futures to maintain (40);
  {\tt NWEIGHTEDFUTURES}, the number of futures that are weighted,
  as opposed to totally random (35);
  {\tt DROPFUTURES}, the number of the worst-scoring futures we
  completely drop (5);
  {\tt MUTATEFUTURES}, the number of worst-scoring futures that
  we replace with mutants of the best future (7);
  {\tt MINFUTURELENGTH} and {\tt MAXFUTURELENGTH}, which put bounds
  on the sizes of futures (50 and 800);
  {\tt OBSERVE\_EVERY}, the frequency with which we sample memory
  for computing the value fraction for motif weighting (10);
  {\tt ALPHA}, the multiplicative factor for up- or down-weighting
  motifs (0.8);
  {\tt MOTIF\_MIN\_FRAC} and {\tt MOTIF\_MAX\_FRAC}, the bounds
  on motif weights (0.1 and 0.00001).

  Each is a scarlet letter of personal shame and future work is to
  eliminate them. In my opinion the least excusable are the bounds on
  future length, since these are related to what portion of time is
  ``interesting'' from a game perspective---for example, the max
  future length must exceed the number of frames that transpire after
  Mario dies but before he respawns, or the algorithm cannot properly
  avoid death. In my opinion this requires too much knowledge of the
  game. I don't have a good idea about how to compute this, at least
  without a human providing an input movie of dying (which would
  help for lots of other reasons)---but that would complicate the
  output of the learning process, which violates a primary design goal.

  {\tt NFUTURES} is another bothersome one, but there is more hope here.
  Basically it trades off computation time for quality of play, and
  more seems to always be better. We do have runtime metrics that could
  be used to dynamically change NFUTURES. For example, we could use
  the notion of improvability from Section~\ref{sec:backtracking}, or
  the value fraction, the gauge the marginal value of searching an
  additional future. Or something like that. This might actually help
  performance because we do the same amount of work (modulo caching)
  even during periods that the inputs are being ignored, like between
  worlds in Super Mario Bros., as we do when the going gets rough,
  and searching for the exact right delicate sequence of inputs would
  benefit from more options. The appearance of pausing in the output
  for Bubble Bobble (Section~\ref{sec:bubblebobble}) suggests that it
  knows all the futures are bad and needs to search different ones,
  and corroborates this idea.}

\subsection{Backtracking} \label{sec:backtracking}

The algorithm above always makes progress at the same rate of 10
frames per iteration. Sometimes it can get stuck in local maxima.
Super Mario Bros., my tuning game for this work, seems almost to have
been designed to thwart {\tt playfun}. The first level is pretty easy
once you have a decent algorithm, and early versions beat it without
any trouble. It's mainly about avoiding Goombas and planning far
enough ahead to make high jumps over pipes and pits (see the
accompanying videos for amusing struggles with these). World {\tt 1-2}
provides a challenge you may recognize from your youth, immediately
followed by something that computers find much harder
(Figure~\ref{fig:mario12}).

I literally spent about six weekends on this problem. First, the
overhang is set up so that enemies emerge from it just as you arrive.
This means that loose play (imperfect enemy avoidance) tends to get
you killed right here. Mario has found a number of different solutions
to this problem, from waiting, to kicking turtles from the left to
clear out the enemies, to timing his jump perfectly (it is possible)
to stomp the turtle after bouncing his head off the ceiling.

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario12}
\end{center}\vspace{-0.1in}
\caption{This is where it starts to get hard, even with lexicographic
ordering and time travel. This overhang is very tight in terms of the
inputs that can get you through it; random play tends to kill you because
of the advancing enemies that can't easily be jumped over. Greedy
algorithms can't resist the ledge with the coins, but then need to
turn around.}
\label{fig:mario12}
\end{figure}

\subsection{Caching of emulation} \label{sec:caching}

\subsection{MARIONET}

\section{Results} \label{sec:results}

@@

\subsection{Super Mario Bros.}

@@

Eventually, Pac-Man gets far enough away from the remaining dots that
none of his futures bring him near, and without any other objective
to seek out, runs into ghosts. He dies on the first level with 13 dots
left.

@@

\subsection{Bubble Bobble} \label{sec:bubblebobble}

@@

I was surprised that automating this game works at all, since there is
no obvious spatial gradient like in Super Mario Bros., and few local
greedy rewards like in Pac-Man. Bub initially wastes his lives (a
common theme in games where the respawn interval is low---it just
looks like a fast way of warping to the corner). When he's near
monsters he shows good tactics in shooting and popping them
simultaneously, even doing so while facing away from them, which I
didn't know was possible. At the beginning of the movie he prefers to
hide in the bottom-left corner, but soon starts jumping all around the
level, bouncing on his own bubbles and so on, almost like a real
player would. He beats more levels than my input movie did! Since it
takes two start button presses to enter the game, the second START is
part of the input motifs. Amusingly, Bub uses this to his advantage to
change the synchronization of his futures, and when the situation gets
really tight, he'll pause the game until a good future comes around
(Figure~\ref{fig:bubblebobble}). Paused states probably look like
modest progress since memory locations like the music cursor are still
advancing.

After a harrowing escape from the ghost thing on level 4, Bub makes it
to level 5 (my input movie quits at the beginning of level 4!), at
which point I stopped the search because it was getting pretty pausey
and I wanted to see some other games before the SIGBOVIK deadline.

@@

\subsection{Tetris}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{tetris}
\end{center}\vspace{-0.1in}
\caption{The joyful noise of objective functions learned for Tetris.
  The first fifth of the movie is navigating menus and looks very
  different from the remainder. There appear to be some frame counters
  identified, which make perfect smooth progress throughout the entire
  movie. I believe discontinuities correspond to placing of pieces;
  some objectives monotonically increase at these points (and are noisy
  in-between), suggesting that they incorporate the score.
}
\label{fig:tetris-objectives}
\end{figure}
